---
title: "Import data from ECDC's Surveillance Atlas of Infectious Diseases"
author: "The Hackathon Team"
date: "29 November 2016"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", message = FALSE)
options(width=100)
```

This tutorial illustrates how one could import csv files manually downloaded from
[ECDC's Surveillance Atlas of Infectious Diseases](http://ecdc.europa.eu/en/data-tools/atlas)
into `R` using the `"sts"` class from the
[**surveillance**](https://CRAN.R-project.org/package=surveillance)
package.
As an example, we have downloaded the monthly number of reported cases of
salmonellosis (serotype AGONA) by all countries, stored in the file
`ECDC_surveillance_data_Salmonellosis.csv`, which we downloaded
manually from the Atlas.


## Import the csv file from ECDC

These data are from *The European Surveillance System - TESSy*, see
[file LICENSE](../LICENSE).

This is how the first 4 lines of the csv file look like:

```{r csvfile}
cat(readLines("data/ECDC_surveillance_data_Salmonellosis.csv", n = 4), sep="\n")
```

The first step is to import the relevant data from the csv file into R.

```{r ecdc}
read_ecdc <- function (file, id = "RegionCode")
{
    ## import csv file
    ecdc <- read.csv(file, header = TRUE, stringsAsFactors = FALSE,
                     na.strings = "-")

    ## exclude aggregate counts for EU and EU/EEA
    ecdc_countries <- subset(ecdc, !RegionName %in% c("EU/EEA", "EU"),
                             select = -TxtValue)

    ## convert from long format to wide format needed by sts
    ecdc_wide <- reshape(ecdc_countries[c("Time", id, "NumValue")],
                         idvar = "Time", timevar = id,
                         direction = "wide")
    names(ecdc_wide) <- sub("NumValue.", "", names(ecdc_wide), fixed = TRUE)
    row.names(ecdc_wide) <- NULL
    attr(ecdc_wide, "reshapeWide") <- NULL
    return(ecdc_wide)
}

ecdc <- read_ecdc("data/ECDC_surveillance_data_Salmonellosis.csv")

head(ecdc)
tail(ecdc)
```


## Import map data

The `"sts"` class can be used without a supplementary map, but incorporating one
enables nice visualizations.
We retrieve a suitable GeoJSON dataset for Europe from the
[GitHub repository of the open knowledge foundation](https://github.com/datasets/geo-nuts-administrative-boundaries).

```{r map_Europe}
library("sp")
read_map_Europe <- function ()
{
    ## Read NUTS1 level data for Europe from open knowledge foundation
    map <- rgdal::readOGR("https://raw.githubusercontent.com/datasets/geo-nuts-administrative-boundaries/master/data/nuts_rg_60m_2010_lvl_1.geojson", "OGRGeoJSON", verbose = FALSE)

    ## omit "FR9" (French Guiana) [no offense] for a more compact map
    map <- map[map$NUTS_ID != "FR9",]

    ## Make a factor variable containing the country
    map$COUNTRY <- factor(substr(map$NUTS_ID,1,2))
    levels(map$COUNTRY)

    ## Union polygons by country
    rgeos::gUnaryUnion(map, id = map$COUNTRY)
}

## if you cannot get rgdal to work on your system, just use the stored file
file_map_Europe <- "data/map_Europe.RData"
if (file.exists(file_map_Europe)) {
    load(file_map_Europe)
} else {
    map_Europe <- read_map_Europe()
    save(map_Europe, file = file_map_Europe, compress = "xz")
}
```

```{r map_summary}
summary(map_Europe)
```

```{r map_plot, out.width = "100%", echo = -1}
par(mar = c(0,0,0,0))
plot(map_Europe)
```

## Load and merge population data

```{r}
population <- read.csv("data/population-eurostat.csv",
                       skip = 11, sep = ";", na.string = ":")
idx <- pmatch(names(ecdc)[-1], population$NUTS)
stopifnot(sum(is.na(idx)) == 0)
year <- paste0("X",unlist(lapply(strsplit(ecdc$Time, split="-"), function(x) x[1])))
popMatrix <- t(population[idx,year])
colnames(popMatrix) <- as.character(population$NUTS)[idx]
```

## Create an `"sts"` object

We can now create an `"sts"` object from the matrix of counts and the map
containing a geographic representation of the columns of the count data.

```{r sts}
library("surveillance")
(start <- as.numeric(strsplit(min(ecdc$Time), split="-")[[1]]))
ecdc_sts <- sts(ecdc[,-1], start = start, frequency = 12, # monthly data
                population = popMatrix,
                map = map_Europe) # row.names(map_Europe) are matched against colnames(ecdc)
```

Now having the data as an object of the `"sts"` class, the
**surveillance** package offers functionality for visualization,
monitoring and modelling of the data. Some basic visualizations are
shown below, the monitoring and modelling features are illustrated in
other tutorials on this site or [vignettes of the surveillance
package](https://cran.r-project.org/web/packages/surveillance/vignettes/hhh4_spacetime.pdf).


`"sts"` objects can also be converted to the well-known
[`"xts"`](https://CRAN.R-project.org/package=xts) class,
to the basic `"ts"` class,
and to a `data.frame` via the associated `as.data.frame()` method.
See [this tutorial](sts-conversions.html).


## Visualizations

### Time series of cases aggregated over all countries

```{r stsplot_time1}
plot(ecdc_sts, type = observed ~ time)  # incomplete because of missing counts
```

### Selecting a single time series

```{r stsplot_time_finland}
plot(ecdc_sts[,"FI"], type = observed ~ time)
```

### Multivariate time series

```{r stsplot_time, fig.width=10, fig.height=7, out.width="100%"}
plot(ecdc_sts, type = observed ~ time | unit)  # see help("stsplot_time") for options
```

Currently, small manual efforts are required to plot the incidence
(a `population` argument as for the spatial plots below is in development):
```{r stsplot_time_incidence, fig.width=10, fig.height=7, out.width="100%"}
local({
    observed(ecdc_sts) <- observed(ecdc_sts) / population(ecdc_sts) * 1000000
    plot(ecdc_sts, type = observed ~ time | unit, ylab = "Incidence (per 1'000'000)")
})
```


### Map aggregated over time

```{r stsplot_space}
plot(ecdc_sts, type = observed ~ unit)  # see help("stsplot_space") for options
```

To change the color key to a color-blind friendly version:

```{r COLS}
(COLS <- viridis::viridis_pal()(9))
```
```{r stsplot_space_colors}
plot(ecdc_sts, type = observed ~ unit, col.regions = COLS)
```


### Incidence map aggregated over time

```{r stsplot_space_incidence}
plot(ecdc_sts, type = observed ~ unit,
     population = (colMeans(population(ecdc_sts))/1000000) * (nrow(ecdc_sts)/12))
     # -> mean yearly incidence per 1'000'000 inhabitants
```


### Animation from 2011 to 2015

```{r animation, fig.show="animate", out.width="75%", fig.width=6, fig.height=8, cache=TRUE}
animate(ecdc_sts, tps = 49:96, sleep = 0, verbose = FALSE,
        timeplot = list(as.Date = TRUE))
## see help("animate.sts") for options ['as.Date' requires surveillance > 1.12.2]
```

The [**animation**](https://CRAN.R-project.org/package=animation) package can be
used to store such an animation in other formats, e.g., Flash, GIF, HTML pages,
PDF and videos. The `help("animate.sts")` contains an example.


### Animation with incidence

```{r animation_incidence, fig.show="animate", out.width="75%", fig.width=6, fig.height=8, cache=TRUE}
animate(ecdc_sts, tps = 49:96, sleep = 0, verbose = FALSE,
        timeplot = list(as.Date = TRUE),
        population = colMeans(population(ecdc_sts))/1000000)
```
