---
title: "Convert `sts` objects to various other classes"
author: "The Hackathon Team"
date: "29 November 2016"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", message = FALSE)
options(width=100)
```

The **surveillance** package holds plenty of data sets from the public health
surveillance of infectious disease occurrence. These count time series
are objects of the class `"sts"` (surveillance time series), as illustrated in,
e.g., the [ECDC tutorial](ECDC2sts.html).

This vignette illustrates how to transform an `"sts"` object (back) to a
`"data.frame"` or an `"xts"` object. We use

```{r stsdata}
library(surveillance)
data("ha.sts")
```


## Data frame

Transforming an `"sts"` object to a data frame `cbind`s the various
contained matrices (observed, state, alarm, upperbound, population):

```{r as.data.frame}
hadata <- as.data.frame(ha.sts, as.Date = TRUE)
head(hadata, 2)
```

This enables post-processing of `sts` objects using other tools.
For instance, to generate a plot of the data using
[**ggplot2**](https://CRAN.R-project.org/package=ggplot2).

First we convert the data into a long format using `tidyr` and `dplyr`.
```{r ggplot2_data, message=FALSE}
library(dplyr)
plot_data <- hadata %>% 
  tidyr::gather(counties, count, starts_with("observed")) %>% 
  tidyr::separate(counties, c("key", "county"), fill = "left") %>% 
  select(epoch, county, count)
head(plot_data)
```

Then it is straight forward to plot data in `ggplot2`.
```{r ggplot2}
library(ggplot2)
ggplot(plot_data, aes(x = epoch, y = count)) +
  geom_bar(stat = "identity", color = "#123456") +
  facet_wrap(~ county) +
  xlab("ISO week") +
  ylab("# cases") 
```


## xts

The `as.xts` function can be used to transform `sts` data to the standard
`"xts"` class for time series (there is also an `as.ts`-method):

```{r as.xts}
library("xts")
haxts <- as.xts(ha.sts)
plot(haxts)
```

Up to old **xts** 0.9-7, this will only plot the first time series
Newer versions of **xts** are available from the
[**xts** GitHub repository](https://github.com/joshuaulrich/xts).

This also enables the use of the
[**dygraphs**](https://CRAN.R-project.org/package=dygraphs)
package for interactive visualization:

```{r dygraph, out.width="100%"}
dygraphs::dygraph(haxts)
```
