---
title: "Monitor the Conference Twitter Feed"
author: "The Hackathon Team"
date: "30 November 2016"
output:
 html_document:
   self_contained: no
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(fig.align = "center", message = FALSE)
options(width=100)
```

This tutorial illustrates how one extracts data from twitter
using `R` in order to quickly establish a tweet based monitoring
system for critical event monitoring.
We will use the `rtweet` package to harvest the data using the
[twitter API](https://dev.twitter.com/overview/api) and subsequently
monitor them with the `surveillance` package.

## Setting up the twitter API

The [README.md](https://github.com/mkearney/rtweet) of the `rtweet`
package provides helpful information on how to create a twitter app to
automatically search tweets using the twitter AP

```{r, message=FALSE}
library(rtweet)
library(ggplot2)
library(dplyr)
```

```{r,eval=FALSE}
# A file containing the information of your twitter app. To protect the
# information stored here, this is kept outside the public git repository.
#
# The contents of the file are simply:
# twitter_token <- create_token(app = "surveillance_trends", # whatever you named app
#   consumer_key = "2zIXXX4L6USa4UfXXXXXXXXXX",
#   consumer_secret = "oXXXXXXXSwwXXXXXXXXXXXXXXsXXXXXXXmXXXXXXXX")

source("~/Sandbox/Twitter-Trends/auth-trends.R",encoding="UTF8")
```

## Performing Queries

Perform the query, here we shall search for tweets containing the
hashtag `#ESCAIDE2016`. The result is a list of individual tweets --
this is almost similar to the well known concept of a linelist:

```{r}
the_query <- "#ESCAIDE2016 OR #ESCAIDE OR @ESCAIDE"
```

```{r,eval=FALSE}
tweets <- search_tweets(the_query, n = 15000, type="recent", token = twitter_token)
```
````{r,echo=FALSE,eval=FALSE}
saveRDS(object=tweets,file="data/tweets.rds")
tweet_query_date <- Sys.time()
dump(list="tweet_query_date",file="data/tweet-query-date.R")
```
```{r,echo=FALSE}
tweets <- readRDS(file="data/tweets.rds")
source(file="data/tweet-query-date.R")
```

A total of `r nrow(tweets)` tweets were collected on `r
as.character(tweet_query_date)`. Below we show the first few entries:

```{r, R.options=list(width=300)}
DT::datatable(head(tweets))
```


### Descriptive Analysis

Who's tweeting? The tweets originate from a total of
`r n_distinct(tweets$screen_name)` users. The top 30 users (by their
number of tweets) are:

```{r,message=FALSE}

top_tweeters <- tweets %>% group_by(screen_name) %>%
  summarise(nTweets=n()) %>%
  arrange(desc(nTweets)) %>%
  top_n(n=30)

ggplot( top_tweeters, aes(x=screen_name,weight=nTweets)) + geom_bar() + coord_flip() + xlab("Number of tweets")
```

Distribution of the different hashtags used in the tweets:

```{r}
df <- data.frame(hashtag=unlist((tweets)$hashtags)) %>%
  mutate(hashtag = stringr::str_trim(hashtag), hashtag = tolower(hashtag)) %>%
  filter(!hashtag %in% c("escaide2016")) %>%
  group_by(hashtag) %>%
  summarise(n=n()) %>% filter(!is.na(hashtag)) %>% arrange(-n) %>% top_n(n=40)

ggplot(df, aes(x=hashtag,y=n)) + geom_bar(stat="identity") +
  theme(axis.text.x=element_text(angle=90, hjust=1))
```

Distribution of the system used for the tweeting:

```{r}
ggplot( tweets, aes(source)) + geom_bar() + coord_flip()
```

Time series of the tweets (without interruptions):
```{r}
library(lubridate)

## Make POSIX function for dplyr re-use
make_posix <- . %>% mutate(time=lubridate::ymd_hms(paste0(year,"-",month,"-",day," ",hour,":00:00"),tz="CET"))

##Are tweets a retweet of our tweet?
#the_tweet_id <- tweets %>% filter(screen_name == "m_hoehle") %>% head(n=1) %>% .[["status_id"]] %>% as.character
tweets <- tweets %>% mutate( is_retweet_of_us = (!is.na(retweet_status_id == 803931871662972928) & retweet_status_id == 803931871662972928))
#tweets2 <- tweets %>% mutate_( paste0("is_retweet_of_us = (retweet_status_id == ", the_tweet_id,")"))

## All possible time points
ts <- tweets %>%
  mutate(hour = sprintf("%.02d",lubridate::hour(created_at)), day=lubridate::day(created_at),month=lubridate::month(created_at),year=lubridate::year(created_at)) %>%
  make_posix %>%
  group_by(time, is_retweet_of_us) %>%
  summarise(n=n(),n_retweets=sum(is_retweet_of_us,na.rm=TRUE),n_noretweets=n-n_retweets)

allTimes <- expand.grid(year=2016, month=11, day=25:30, hour=sprintf("%.02d",0:23)) %>%
  make_posix %>%
  arrange(time) %>% mutate(n=0)

ts2 <- allTimes %>% left_join(ts, by="time") %>% group_by(time) %>%
  filter(time > "2016-11-28 08:00:00 CET" & (time <= "2016-11-30 19:00:00 CET")) %>% select(-n.x,-is_retweet_of_us,-n.y)

na2zero <- function(x) {x[is.na(x)] <- 0 ; return(x)}

ts2 <- ts2 %>% mutate(n_retweets = na2zero(n_retweets),
                      n_noretweets = na2zero(n_noretweets)) %>%
  rename(retweet = n_retweets, no_retweet = n_noretweets)
ts3 <- tidyr::gather(ts2, group, n, retweet:no_retweet)
ts3 <- ts3 %>% mutate(group = factor(group))

```


```{r}
p <- ggplot(ts3) +
  geom_bar(aes(x=time, y=n, fill=group), stat="identity",alpha=.8) +
  theme(axis.text.x=element_text(angle=90, hjust=1)) +
  ylab("No. of tweets") +
  xlab("Time") + ggtitle(the_query) +
  ##  scale_x_datetime(date_minor_breaks="1 hour",date_breaks="1 hour",date_labels="Nov %d-%H:00") +
  ## http://stackoverflow.com/questions/36227130/r-as-posixct-timezone-and-scale-x-datetime-issues-in-my-dataset
  scale_x_datetime(date_minor_breaks="1 hour",date_breaks="1 hour",labels=scales::date_format("Nov %d-%H:00", tz = "CET"),expand=c(0,0)) +
  theme(axis.text.x = element_text(angle = 60, size = 6), legend.position = "bottom") +
  viridis::scale_fill_viridis(discrete = TRUE, name="Tweet Type", label = c("Standard tweet",
                                                                            "Outbreak tweet/retweet"))
p
```

Now we overlay the conference program

```{r}
conference_program_slots <- tibble::tribble(
  ~from, ~to, ~slot_type,
  "2016-11-28 09:00:00 CET",   "2016-11-28 10:30:00 CET", "Plenary Talks",
  "2016-11-28 11:00:00 CET",   "2016-11-28 12:40:00 CET", "Parallel Session",
  "2016-11-28 14:30:00 CET",   "2016-11-28 15:30:00 CET", "Parallel Session",
  "2016-11-28 15:30:00 CET",   "2016-11-28 16:30:00 CET", "Poster Session",
  "2016-11-28 17:00:00 CET",   "2016-11-28 18:30:00 CET", "Plenary Talks",

  "2016-11-29 09:00:00 CET",   "2016-11-29 10:30:00 CET", "Plenary Talks",
  "2016-11-29 11:00:00 CET",   "2016-11-29 12:40:00 CET", "Parallel Session",
  "2016-11-29 14:30:00 CET",   "2016-11-29 15:30:00 CET", "Parallel Session",
  "2016-11-29 15:30:00 CET",   "2016-11-29 16:30:00 CET", "Poster Session",
  "2016-11-28 17:00:00 CET",   "2016-11-28 18:30:00 CET", "Parallel Session",

  "2016-11-30 09:00:00 CET",   "2016-11-30 10:30:00 CET", "Plenary Talks",
  "2016-11-30 11:00:00 CET",   "2016-11-30 12:40:00 CET", "Parallel Session",
  "2016-11-30 14:30:00 CET",   "2016-11-30 15:30:00 CET", "Parallel Session",
  "2016-11-30 15:30:00 CET",   "2016-11-30 16:30:00 CET", "Poster Session",
  "2016-11-30 17:00:00 CET",   "2016-11-30 18:30:00 CET", "Plenary Talks"
) %>% mutate(from = as.POSIXct(from), to = as.POSIXct(to))
```

```{r}
p + geom_rect(data = filter(conference_program_slots, to <= "2016-11-30 19:00:00 CET"),
            aes(xmin = from, xmax = to, ymin = 0, ymax = max(ts3$n), color = slot_type), alpha = 0.4) +
  theme(legend.position = "bottom") +
  viridis::scale_color_viridis(discrete = TRUE, name = "Session Type:")
```

## Outbreak detection

We shall use the EARS C method for performing outbreak detection.

```{r ears}
library(surveillance)
baseline <- 7
tweet_sts <- surveillance::sts(observed = ts2$retweet + ts2$no_retweet, # weekly number of cases
                               epoch = as.numeric(ts2$time))

monitored_tweets <- earsC(tweet_sts, control = list(baseline = baseline))
monitored_tweets_df <- as.data.frame(monitored_tweets)

monitored_tweets_df <- mutate(monitored_tweets_df,
                          time = ts2$time[(baseline + 1):(nrow(ts2))])
ggplot(ts3) +
  geom_bar(aes(time, n, fill = group), stat = "identity") +
  viridis::scale_fill_viridis(discrete = TRUE, name = "Alarm:") +
  geom_step(data = monitored_tweets_df, aes(time, upperbound)) +
  theme(legend.position = "bottom") +
  theme(axis.text.x=element_text(angle=90, hjust=1)) +
  ylab("No. of tweets") +
  xlab("Time") + ggtitle(the_query) +
    ##  scale_x_datetime(date_minor_breaks="1 hour",date_breaks="1 hour",date_labels="Nov %d-%H:00") +
  scale_x_datetime(date_minor_breaks="1 hour",date_breaks="1 hour",labels=scales::date_format("Nov %d-%H:00", tz = "CET"),expand=c(0,0)) +
  theme(axis.text.x = element_text(angle = 60, size = 6)) +
  geom_point(data = filter(monitored_tweets_df, alarm), aes(x = time), y = 0, color = "red") +
  viridis::scale_fill_viridis(discrete = TRUE, name="Tweet Type", label = c("Standard tweet",
                                                                            "Our outbreak tweet/retweet"))

```

### Instant citizen science experiment

Let's see, if we can artificially inject an outbreak into the time
series. Retweet the post below, if you want to participate in an
instant citizen science experiment.

<center>
<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Syndromic surveillance at <a href="https://twitter.com/hashtag/escaide2016?src=hash">#escaide2016</a> ! Let&#39;s retweet &amp; create a virtual outbreak 😷 <a href="https://t.co/zks5PZOTpM">https://t.co/zks5PZOTpM</a> <a href="https://twitter.com/ma_salmon">@ma_salmon</a> <a href="https://twitter.com/dirk_sch">@dirk_sch</a> <a href="https://twitter.com/hashtag/rstats?src=hash">#rstats</a> <a href="https://t.co/TLzm3oCvDM">pic.twitter.com/TLzm3oCvDM</a></p>&mdash; Michael Höhle (\@m_hoehle) <a href="https://twitter.com/m_hoehle/status/803931871662972928">November 30, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js"
charset="utf-8"></script>
</center>

For a not so serious version of the above citizen science experiment,
see this
[blog post](http://staff.math.su.se/hoehle/blog/2016/09/25/sootb.html)
on how to detect zombie outbreaks using twitter.

